---
title: "info201-final"
author: "Erika Cui, Varsha Bharath, Nithika Yechuri"
date: "2023-12-06"
output: html_document
---

```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(stringr)
```

#### **Data Sources**

Here is the first source we used: https://www.kaggle.com/datasets/paultimothymooney/zillow-house-price-data?selhttps%3A%2F%2Fwww.kaggle.com%2Fdatasets%2Fpaultimothymooney%2Fzillow-house-price-data%3Fselect=Sale_Prices_State.csvect%3DSale_Prices_State.csv&select=Sale_Prices_State.csv

The second data: https://www.kaggle.com/datasets/kabhishm/united-states-counties-by-per-capita-income

The third data: https://www.kaggle.com/datasets/neerajkld/realtor-real-estate-usa 

#### **Variables**

states: State name

price: Price of the house

SizeRank: The ranking of the state, depending on the size of region

avg_price: Average price of a house in a certain state, after collecting all the price from each city in the               state

price_category: Shows if the price of the house is considered "High", "Medium", or "Low"

avg_pci: The average per capita income for each state

#### **Load, Clean, Average**

We first load the dataset. Then we select, the variables that were the most informational, and renamed them.

```{r}
salePriceData <- read_delim("../data/Sale_Prices_State.csv")

salePrice <- salePriceData %>% 
  select(RegionName, `2020-02`, SizeRank) %>% 
  rename(states = RegionName, price = `2020-02`)

head(salePrice) 

dim(salePrice)
missing <- colSums(is.na(salePrice) | salePrice == "NA")
print(missing)
```
Summary: There are 49 rows, 3 columns, 0 missing values.


Here is the plot of the data set. Organized by least to most pricey houses in the different states.

```{r}
salePrice %>% 
  ggplot(aes(x = reorder(states, -price), y = price, fill = states)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme(axis.text.y = element_text(size = 6)) +
  theme(legend.text = element_text(size = 7),
        legend.title = element_text(size = 10),
        legend.key.size = unit(0.5, "cm")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "House Price in Different States", x = "States", y = "Price")
```

We load the second data, selecting the most useful columns.

```{r}
countiesData <- read_delim("../data/counties_per_capita_income.csv")

pciData <- countiesData %>% 
  select(states, pci)
head(pciData)

dim(pciData)
missing <- colSums(is.na(pciData) | pciData == "NA")
print(missing)
```
Summary: 3231 rows, 2 columns, and 0 missing values.

Piping the pciData down, the graph will show us the pattern between states and the pci. 

```{r}
repeatStates <- pciData %>%
  select(states) %>% 
  distinct() %>% 
  pull()

pciData <- pciData %>%
  mutate(pci = as.numeric(gsub("[$,]","", pci))) %>% 
  filter(!is.na(pci)) %>% 
  arrange(desc(pci)) %>% 
  filter(!duplicated(states)) %>% 
  mutate(states = factor(states, levels = repeatStates))

pciData %>% 
ggplot(aes(x = states, y = pci, fill = states)) +
  geom_bar(stat = "identity", fill = "lightslateblue", width = 0.8) +
  coord_flip() +
  theme(axis.text.y = element_text(size = 6)) +
  theme(legend.text = element_text(size = 7),
        legend.title = element_text(size = 10),
        legend.key.size = unit(0.5, "cm")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "States Per Capita Income", x = "States", y = "Per Capita Income")
```

This is the last data set we will use. Selecting the variables "states" (as that is the merge key) and "price".

```{r}
reData <- read_delim("../data/realtor-data.zip.csv.zip")

realEstate <- reData %>%
  rename(states = state, price = price) %>% 
  select(states, price) %>%
  filter(states != "Puerto Rico")

head(realEstate)

dim(realEstate)
missing <- colSums(is.na(realEstate) | realEstate == "NA")
print(missing)
```
Summary: 1179387 rows, 2 columns, 77 missing values in the column "price".

There are many rows because this data includes multiple cities from each state. To simplify everything, we took all the cities in a state and averaged it out, taking that average as the average price of a house in that certain state.
```{r}
realEstateAvg <- realEstate %>% 
  group_by(states) %>% 
  summarize(avg_price = mean(price)) %>% 
  filter(!is.na(avg_price))
realEstateAvg

dim(realEstateAvg)
missing <- colSums(is.na(realEstateAvg) | realEstateAvg == "NA")
print(missing)
```
Summary: 14 rows, 2 columns, and 0 missing values.


This dataset only has a selected amount of states, so this bar graph shows the average price of a house in those states, giving a visual representation.
```{r}
realEstateAvg %>% 
  ggplot(aes(x = reorder(states, avg_price), y = avg_price)) +
  geom_bar(stat = "identity", fill = "cyan4") +
  coord_flip() +
  labs(title = "Average House Prices in Selected States", x = "States", y = "Average Price")
```

#### **Merge**

```{r}
merge1 <- full_join(salePrice, realEstateAvg, by = "states")

merge1$avg_price[is.na(merge1$avg_price)] <- merge1$price[is.na(merge1$avg_price)]

merge1$avg_price[!is.na(merge1$price) & !is.na(merge1$avg_price)] <- 
  (merge1$price[!is.na(merge1$price) & !is.na(merge1$avg_price)] +
  merge1$avg_price[!is.na(merge1$price) & !is.na(merge1$avg_price)]) / 2

merge1 <- merge1 %>% 
  arrange(states) %>%
  select(states, avg_price) %>% 
  arrange(desc(avg_price)) %>%
  filter(states != "American Samoa") %>% 
  filter(states != "Puerto Rico") %>% 
  filter(states != "Guam") %>% 
  filter(states != "Northern Mariana Islands") %>% 
  filter(states != "U.S. Virgin Islands") %>% 
  filter(states != "Virgin Islands")

head(merge1)

dim(merge1)
missing <- colSums(is.na(merge1) | merge1 == "NA")
print(missing)
```
Summary: 50 rows, 3 columns, 0 missing values.


With the merged data, some states in one data may not have an average price. To fix this, if it is detected that a state has a price but no average, it takes the price and moves it to the average. If the state has both a price and an average, it averages those two and sets the number to avg_price.

```{r}
merge1 %>%
  mutate(states = factor(states, levels = states[order(avg_price)])) %>% 
  ggplot(aes(x = states, y = avg_price)) +
  geom_bar(stat = "identity", fill = "springgreen3", width = 0.8) +
  coord_flip() +
  theme(axis.text.y = element_text(size = 6)) +
  theme(legend.text = element_text(size = 7),
        legend.title = element_text(size = 10),
        legend.key.size = unit(0.5, "cm")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Average Price for Houses Across the U.S", x = "States", y  = "Average Price")
```
```{r}
merge1 <- merge1 %>% 
  mutate(price_category  = ifelse(avg_price < 200000, "Low",
                               ifelse(avg_price < 400000, "Medium", "High")))
head(merge1)

dim(merge1)
missing <- colSums(is.na(merge1) | merge1 == "NA")
print(missing)
```
Summary:50 rows, 3 columns, 0 missing values.


```{r}
merge2 <- left_join(pciData, merge1, by = "states")

merge2$pci <- as.numeric(gsub("[^0-9.]", "", merge2$pci))

merge2 <- merge2 %>% 
  group_by(states) %>% 
  summarize(avg_pci = mean(pci, na.rm = TRUE)) %>% 
  filter(states != "American Samoa") %>% 
  filter(states != "Puerto Rico") %>% 
  filter(states != "Guam") %>% 
  filter(states != "Northern Mariana Islands") %>% 
  filter(states != "U.S. Virgin Islands") %>% 
  filter(states != "Virgin Islands")

head(merge2)

dim(merge2)
missing <- colSums(is.na(merge2) | merge2 == "NA")
print(missing)
```
Summary: 50 rows, 2 columns, 0 missing values.


```{r}
merge2 %>%
  mutate(states = factor(states, levels = states[order(avg_pci)])) %>% 
  ggplot(aes(x = states, y = avg_pci)) +
  geom_bar(stat = "identity", fill = "honeydew3", width = 0.8) +
  coord_flip() +
  theme(axis.text.y = element_text(size = 6)) +
  theme(legend.text = element_text(size = 7),
        legend.title = element_text(size = 10),
        legend.key.size = unit(0.5, "cm")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Average Per Capita Income in the States", x = "States", y  = "Average PCI")
```


```{r}
merge3 <- full_join(merge1, merge2, by = "states")

merge3 <- merge3 %>% 
  filter(states != "Virgin Islands")

head(merge3)

dim(merge3)
missing <- colSums(is.na(merge3) | merge3 == "NA")
print(missing)
```

Summary: 51 rows, 5 columns, 1 missing value in "avg_price", 1 missing value in "price_category", and 1 missing 
value in "avg_pci".


#### **Story Line Graphs**

```{r}
merge3 %>% 
  ggplot(aes(x = avg_pci, y = avg_price, label = states)) +
  geom_point() +
  geom_text_repel(size = 3, force = 10) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Minimal Income to Afford a House in Different States", x = "Average PCI", y = "Average Price")
```

```{r}
top10 <- merge3 %>% 
  arrange(desc(avg_price)) %>% 
  slice(1:10)

ggplot(top10, (aes(x = reorder(states, -avg_price), y = avg_price))) +
  geom_bar(stat = "identity", fill = "purple2")+
  scale_y_continuous(labels = scales::comma) +
  coord_flip() +
  labs(title = "Top 10 Most Expensive States to Buy a House", x = "Average Price", y = "States")
```

```{r}
top5 <- merge3 %>% 
  arrange(desc(avg_price)) %>% 
  slice(1:5)
```

```{r}
bot5 <- merge3 %>% 
  arrange(avg_price) %>% 
  slice(1:5)
```

```{r}
top5$group <- "Top 5"
bot5$group <- "Bottom 5"

mergeGroup <- rbind(top5, bot5)

ggplot(mergeGroup, aes(x = reorder(states, avg_price), y = avg_price, fill = group)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Top 5" = "violetred", "Bottom 5" = "lightblue"), guide = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  coord_flip() +
  theme(axis.text.y = element_text(size = 6)) +
  labs(title = "Income Difference Between Top 5 and Bottom 5", x = "States", y = "Average Price")
```


```{r}
regions <- function(states) {
  ifelse(states %in% c("Washington", "Orgeon", "California", "Nevada"), "West Coast",
         ifelse(states %in% c("Montana", "Idaho", "Wyoming", "Utah", "Colorado", "Arizona", "New Mexico"), "Mountain Coast",
                ifelse(states %in% c("North Dakota", "South Dakota", "Nebraska", "Kansas", "Oklahoma", "Texas", "Minnesota", "Iowa", "Missouri", "Arkansas", "Louisiana", "Wisconsin", "Illinois", "Tennessee", "Mississippi", "Alabama", "Kentucky"), "Midwest Coast",
                       ifelse(states %in% c("Michigan", "Indiana", "Ohio", "West Virginia", "Virginia", "Georgia", "Florida", "North Carolina", "South Carolina", "Pennsylvania", "New York", "Massachusetts", "Vermont", "New Hampshire", "Maine", "Rhode Island", "Conneticut", "New Jersey", "Delaware", "Maryland", "District of Columbia"), "East Coast", "Other"))))
}

merge3 <- merge3 %>% 
  mutate(region = regions(states))

regionData <- merge3 %>% 
  group_by(region) %>% 
  summarize(avg_price = mean(avg_price, na.rm = TRUE))

regionData %>% 
ggplot(aes(x = region, y = avg_price, color = region)) +
  geom_bar(stat = "identity", fill = "azure2", width = 0.5) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Average Price in Different Coasts", x = "Region", y = "Average Price")
  
```

#### **Summarization Table**

```{r}
finalData <- merge3 %>% 
  select(states, avg_pci, avg_price, price_category)

head(finalData)

dim(finalData)
missing <- colSums(is.na(finalData) | finalData == "NA")
print(missing)
```
Summary: 51 columns, 4 columns, 1 missing value in "avg_pci", 1 missing value in "avg_price", and 1 missing value in "price_category".
